Short-time simulation of open collective quantum spin systems

### 2018-12-03 soc_squeezing/paper/collective_spin.tex
removed obsolete text
modified first intro paragraph, added some references

Collective spin systems attract considerable attention for providing a
window into many-body quantum phenomena that are both experimentally
accessible\cite{takano2009spin, chen2010heralded,
  riedel2010atomchipbased, gross2010nonlinear, martin2013quantum,
  bohnet2016quantum, hosten2016quantum, norcia2018cavitymediated,
  jones2009magnetic} and metrologically
useful\cite{riedel2010atomchipbased, gross2010nonlinear,
  hosten2016quantum, jones2009magnetic, zhong2010simplified,
  huang2015quantum, agarwal1997atomic, lau2014proposal,
  wineland1992spin, kitagawa1993squeezed, ma2011quantum}.  Theoretical
interest in such systems dates back to the mid-twentieth century with
the introduction of the Lipkin-Meshkov-Glick (LMG) model as a toy
model for testing many-body approximation methods in contemporary
nuclear physics\cite{lipkin1965validity, meshkov1965validity,
  glick1965validity}.  Since then, collective system systems have
additionally featured in theoretical studies of quantum
criticality\cite{latorre2005entanglement, alcalde2007functional,
  morrison2008collective, sarandy2009classical, wang2012quantum,
  majd2014lmg}, non-equilibrium
phenomena\cite{morrison2008dissipationdriven,
  bhattacherjee2014nonequilibrium, klinder2015dynamical,
  zhiqiang2017nonequilibrium, smale2018observation}, and precision
metrology\cite{wineland1992spin, kitagawa1993squeezed, ma2011quantum,
  zhong2010simplified}.  With the development of advanced atomic
trapping, cooling, and control techniques, collective spin systems
have been experimentally realized in atomic
ensembles\cite{takano2009spin, chen2010heralded}, Bose-Einstein
condensates\cite{riedel2010atomchipbased, gross2010nonlinear},
ultracold Fermi gasses\cite{martin2013quantum}, trapped
ions\cite{bohnet2016quantum}, optical cavities\cite{hosten2016quantum,
  norcia2018cavitymediated}, and molecular
sensors\cite{jones2009magnetic}.

One of the primary motivations for studying collective spin systems is
their application to quantum-enhanced metrology.  Quantum projection
noise limits the error $\Delta\phi$ in the measurement of a phase
angle $\phi$ with $N$ independent spins to
$\Delta\phi\sim1/\sqrt{N}$\cite{wineland1992spin, itano1993quantum,
  ma2011quantum}.  Collective spin systems enable a means to break
through this limit via the preparation of many-body entangled states
such as NOON\cite{jones2009magnetic, chen2010heralded,
  zhong2010simplified}, spin-cat\cite{huang2015quantum,
  agarwal1997atomic, lau2014proposal}, and spin-squeezed
states\cite{takano2009spin, riedel2010atomchipbased,
  gross2010nonlinear, hosten2016quantum, norcia2018cavitymediated,
  wineland1992spin, kitagawa1993squeezed, ma2011quantum} that allow
for measurement errors $\Delta\phi\sim1/N^\epsilon$ with
$\epsilon>1/2$.  Generating such states requires nonlinear dynamics
that are typically realized using long-range collisional, photon-, or
phonon-mediated interactions.
% typically resulting in a one-axis twising (OAT) model described by
% the Hamiltonian\cite{kitagawa1993squeezed}
% \begin{align}
%   H_{\t{OAT}} = \chi S_\z^2,
%   \label{eq:OAT}
% \end{align}
% where $S_\z\equiv\sum_{j=1}^N\sigma_\z^{(j)}/2$ is a collective
% spin-$z$ operator defined in terms of Pauli-$z$ operators
% $\sigma_\z^{(j)}$ addressing spin $j$.  The OAT model prepares a
% spin-squeezed state that allows measurement of a phase angle $\phi$
% with error $\Delta\phi\sim N^{-2/3}$ at times
% $t\sim\chi^{-1}N^{-2/3}$\cite{ma2011quantum}.
Although a truly collective spin model requires uniform, all-to-all
interactions, as long as measurements do not distinguish between
constituent particles even non-uniform systems can be effectively
described by a uniform model with renormalized
parameters\cite{hu2015entangled}.
% Various works have proposed improvements over OAT through the
% addition of auxiliary control fields\cite{law2001coherent,
% liu2011spin, huang2015twoaxis, muessel2015twistandturn} or a
% modified vacuum\cite{hu2017vacuum}, thereby realizing different
% collective spin models.

% While the OAT model admits exact
% solutions\cite{foss-feig2013nonequilibrium}, the same is not true of
% more general, non-Ising Hamiltonians.  Nonetheless,
In the absence of decoherence, permutation symmetry and total spin
conservation divide the total Hilbert space of a collective spin
system into superseletion sectors that grow only linearly with system
size $N$, thereby admitting efficient classical simulation of its
dynamics.  Decoherence will generally violate total spin conservation
and require the use of density operators, increasing the dimension of
accessible state space to $\sim N^3$\cite{hartmann2016generalized,
  xu2013simulating} and allowing for exact simulations of
$N\lesssim100$ particles.  If decoherence is sufficiently weak,
dynamics are numerically solvable for $N\lesssim10^4$ particles via
``quantum trajectory'' Monte Carlo methods\cite{plenio1998quantumjump,
  zhang2018montecarlo} (also known as ``quantum jump'' or ``Monte
Carlo wavefunction'' methods) that can reproduce all expectation
values of interest.  When decoherence is strong, however, or when the
number of jump operators grows extensively with system size (e.g.~for
single-spin decay), these Monte Carlo methods can take a prohibitively
long time to converge, as simulations become dominated by incoherent
jumps that generate large numbers of distinct trajectories over which
to average.

@@

In principle, the factorial suppression of terms at higher orders
(i.e.~in $k$) of the expansion in Eq.~\eqref{eq:order_truncation}
implies for any given time $t$, there exists a truncation order $M_t$
for which truncation error is negligibly small.  In practice, only a
fixed maximal truncation order $\overline M$ is accessible with
limited computational resources, such that
Eq.~\eqref{eq:order_truncation} only allows computation of the
correlator $\bk{\S_{\v m}}$ to a maximal time $t_{\overline M}$.  The
maximal accessible truncation order $\overline M$ generally depends on
the choice of operator basis for collective spin systems, which we
discuss briefly at the end of Appendix \ref{sec:new_basis}.  As we
will see in the following sections, our choice of the basis
$\set{\S_{\v m}}$ allows us to accurately compute spin correlators to
times that are relevant e.g.~for spin squeezing protocols.

>>>>

Collective spin systems are a versatile resource in quantum science
for a range of applications including quantum-enhanced metrology,
quantum simulation, and quantum information.  Theoretical interest in
such systems dates back to the mid-twentieth century with the
introduction of the Lipkin-Meshkov-Glick (LMG) model as a toy model
for testing many-body approximation methods in contemporary nuclear
physics\cite{lipkin1965validity, meshkov1965validity,
  glick1965validity}.  Since then, collective system systems have
additionally featured in both theoretical and experimental studies of
quantum criticality\cite{latorre2005entanglement,
  alcalde2007functional, morrison2008dissipationdriven,
  morrison2008collective, sarandy2009classical, wang2012quantum,
  kessler2012dissipative, majd2014lmg}, non-equilibrium
phenomena\cite{walls1978nonequilibrium, morrison2008dynamical,
  bhattacherjee2014nonequilibrium, klinder2015dynamical,
  maghrebi2016nonequilibrium, zhiqiang2017nonequilibrium,
  lang2018concurrence, smale2018observation}, and precision
metrology\cite{wineland1992spin, kitagawa1993squeezed, ma2011quantum,
  zhong2010simplified}.  With the development of advanced atomic
trapping, cooling, and control techniques, collective spin systems
have been experimentally realized in atomic
ensembles\cite{takano2009spin, chen2010heralded}, Bose-Einstein
condensates\cite{riedel2010atomchipbased, gross2010nonlinear,
  klinder2015dynamical}, ultracold Fermi
gasses\cite{martin2013quantum, smale2018observation}, trapped
ions\cite{bohnet2016quantum}, optical cavities\cite{hosten2016quantum,
  zhiqiang2017nonequilibrium, lewis-swan2018robust,
  norcia2018cavitymediated}, and molecular
sensors\cite{jones2009magnetic}.

One of the primary motivations for studying collective spin systems is
their application to quantum-enhanced metrology.  Quantum projection
noise limits the error $\Delta\phi$ in the measurement of a phase
angle $\phi$ with $N$ independent spins to
$\Delta\phi\sim1/\sqrt{N}$\cite{wineland1992spin, itano1993quantum,
  ma2011quantum}.  Collective spin systems enable a means to break
through this limit via the preparation of many-body entangled states
such as NOON\cite{jones2009magnetic, chen2010heralded,
  zhong2010simplified}, spin-cat\cite{agarwal1997atomic,
  lau2014proposal, huang2015quantum}, and spin-squeezed
states\cite{takano2009spin, riedel2010atomchipbased,
  gross2010nonlinear, hosten2016quantum, lewis-swan2018robust,
  norcia2018cavitymediated, wineland1992spin, kitagawa1993squeezed,
  ma2011quantum} that allow for measurement errors
$\Delta\phi\sim1/N^\epsilon$ with $\epsilon>1/2$.  Generating such
states requires nonlinear dynamics that are typically realized using
long-range collisional, photon-, or phonon-mediated interactions.
Although a truly collective spin model requires uniform, all-to-all
interactions, as long as measurements do not distinguish between
constituent particles even non-uniform systems can be effectively
described by a uniform model with renormalized
parameters\cite{hu2015entangled}.

In the absence of decoherence, permutation symmetry and total spin
conservation divide the total Hilbert space of a collective spin
system into superseletion sectors that grow only linearly with system
size $N$, thereby admitting efficient classical simulation of its
dynamics.  Decoherence will generally violate total spin conservation
and require the use of density operators, increasing the dimension of
accessible state space to $\sim N^3$\cite{hartmann2016generalized,
  xu2013simulating} and allowing for exact simulations of
$N\lesssim100$ particles.  If decoherence is sufficiently weak,
dynamics can be numerically solvable for $N\lesssim10^5$ particles via
``quantum trajectory'' Monte Carlo methods\cite{plenio1998quantumjump,
  zhang2018montecarlo} (also known as ``quantum jump'' or ``Monte
Carlo wavefunction'' methods) that can reproduce all expectation
values of interest.  When decoherence is strong, however, or when the
number of jump operators grows extensively with system size (e.g.~for
single-spin decay), these Monte Carlo methods can take a prohibitively
long time to converge, as simulations become dominated by incoherent
jumps that generate large numbers of distinct trajectories over which
to average.

@@

In principle, the factorial suppression of terms at higher orders
(i.e.~in $k$) of the expansion in Eq.~\eqref{eq:order_truncation}
implies for any given time $t$, there exists a truncation order $M_t$
for which truncation error is negligibly small.  In practice, only a
fixed maximal truncation order $\overline M$ is accessible with
limited computational resources, such that
Eq.~\eqref{eq:order_truncation} only allows computation of the
correlator $\bk{\S_{\v m}}$ to a maximal time $t_{\overline M}$.  The
maximal accessible truncation order $\overline M$ generally depends on
the choice of operator basis for collective spin systems, which we
discuss briefly at the end of Appendix \ref{sec:new_basis}.  As we
will see in the following sections, our choice of basis elements
$\S_{\v m}=S_+^{m_+} S_\z^{m_\z} S_-^{m_-}$ allows us to accurately
compute spin correlators to times that are relevant e.g.~for spin
squeezing protocols.
